---
title: "multiTEMPTED Vignette"
output: github_document
---

If working in Docker container, run the following code...
```{zsh, eval=FALSE}
# emulates amd64 if working on non-linux and builds container
docker build . --platform=linux/amd64 -t multi.tempted

# ports to 8787 for rocker container
docker run \
  --platform=linux/amd64 \
  -v $(pwd):/home/rstudio/work \
  -e PASSWORD=123 \
  -p 8787:8787 \
  multi.tempted
```

Then in containerized R-studio session,\n
1. USER: rstudio, PS: 123 \n
2. open File > Open Project then select multi.tempted.Rproj (adds git tab)\n
3. once opened, run...
```{r, eval=FALSE}
setwd('~/work')
library(devtools)
library(testthat)
library(roxygen2)
library(knitr)
```

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# multiTEMPTED

<!-- badges: start -->
<!-- badges: end -->

The goal of multiTEMPTED is to to implement the statistical method TEMPoral TEnsor Decomposition (TEMPTED) generalized to multiple modalities.

Package dependencies: R (>= 4.5.3), np (>= 0.60-17), ggplot2 (>= 3.4.0), methods (>= 4.2.1)

Run time ____?____

You can cite this paper for using TEMPTED: ____?____

The statistical theories behind TEMPTED can be found in this paper:

____?____

____?____

## Installation

You can install the development version of multi.tempted from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("loulind/multi.tempted")
```

____?_____

## Load packages for this vignette
## Read the example data
## Running TEMPTED for different formats of data
## Run TEMPTED for Microbiome Count Data (Straightforward Way)
### ___?steps?____
## Run TEMPTED for Microbiome Compositional Data (Straightforward Way)
### ___?steps?____
## Run TEMPTED for General Form of Multivariate Longitudinal Data (Straightforward Way)
### ___?steps?____
## Run TEMPTED in Customized Way
### ___?steps?____
## Transferring TEMPTED result from training to testing data
### ___?steps?____
