---
title: "multiTEMPTED Vignette"
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

## Introduction of multiTEMPTED

`multi.tempted` extends TEMPTED (TEMPoral TEnsor Decomposition) to several
longitudinal modalities at once, jointly decomposing M subject × feature × time
tensors into:

- **A** (subjects × r) — shared subject scores; the thread tying the modalities together.
- **B_m** (features × r) — per-modality feature loadings.
- **Zeta_m** — per-modality smooth temporal curves.
- **Lambda** (M × r) — per-modality magnitude of each component.

Modalities may differ in their features, feature counts, and sampling times.

**Citations:** TEMPTED, the single-modality precursor
([Shi et al.](https://doi.org/10.1101/2023.07.26.550749)); and the underlying
theory ([Han et al.](https://doi.org/10.1080/01621459.2022.2153689)).

---

## Installation

Install the development version from
[GitHub](https://github.com/loulind/multi.tempted) with either helper:

```{r install, eval=FALSE}
# with pak (preferred)
pak::pkg_install("loulind/multi.tempted")

# or with remotes (depreciated) remotes::install_github("loulind/multi.tempted")
```
---

## Load packages

```{r load_library, message=FALSE, warning=FALSE}
library(multi.tempted)

# extras used only for visualisation below
library(tidyverse)
library(corrplot)
library(plotly)
library(igraph)
library(ggraph)
library(tidygraph)
library(patchwork)
library(magick)
library(pheatmap)
library(np)
```

---

## Read the example data

The [iPOP exercise study](https://med.stanford.edu/snyderlab/ipop.html): four
omic modalities (**cytokine**, **metabolome**, **lipid**, **protein**), each 
provided preprocessed on a log₁₀ scale, measured at baseline (preworkout) 
and four time points post-workout. `ipop` also carries subject metadata (`meta_subj`) 
and per-sample `SubjectID` / `timepoint` (`meta_visit`). PCA was also performed 
on the lipid data to reduce its dimensionality for the purpose of this vignette. 

```{r load_data, message=FALSE, warning=FALSE}
ipop <- multi.tempted::ipop

names(ipop)
# [1] "meta_subj" "meta_visit" "cytokine" "metabolome" "lipid" "protein"
```

```{r prep_data, message=FALSE, warning=FALSE}
# Stack the four modality tables into a named list
featuretables <- lapply(3:6, function(m) as.matrix(ipop[[m]]))
names(featuretables) <- names(ipop)[3:6]
M <- length(featuretables)

# Map visit codes to approximate minutes from baseline
timepoint <- as.vector(ipop[[2]]$timepoint)
timepoint <- case_when(
  timepoint == 1 ~ 0,
  timepoint == 2 ~ 12,
  timepoint == 3 ~ 25,
  timepoint == 4 ~ 40,
  timepoint == 5 ~ 70
)
timepoints <- rep(list(timepoint), M)
subjectID  <- rep(list(as.character(ipop[[2]]$SubjectID)), M)

# Grouping: a per-sample sex vector, and a subjectID--sex table
group_sex   <- ipop[[1]]$Sex
group_subID <- ipop[[1]][, c("subjectID", "Sex")]
```

---

## Running multiTEMPTED

### Run multiTEMPTED

`multitempted_all()` runs the whole pipeline; see `?multitempted_all` for every
argument. The decisions that matter most:

- `transforms`: `"clr"` for raw microbiome counts, `"none"` for already-scaled data.
- `r`: number of components. `centralize`: remove each modality's mean structure first (default `TRUE`).
- `do_ratio`: log-ratio meta-features — `TRUE` only for raw counts.
- `weights`: down-weight a dominant modality when scales differ (see note below).

> **Signs** of A, B, and Zeta can flip together, exactly as in SVD — read
> relative patterns, not absolute sign.

> **Scale:** the shared A pools least-squares signal, so a larger-scale modality
> pulls harder. These modalities share a log₁₀ scale, so equal weights are fine;
> otherwise standardise first or set `weights` ∝ 1 / modality variance.

```{r run_multitempted}
output <- multitempted_all(
  featuretables = featuretables,
  timepoints    = timepoints,
  subjectID     = subjectID,
  transforms    = "none", # data are already log10-transformed
  do_ratio      = FALSE,  # not raw counts
  r             = 3 # number computed components (will compute more later)
)
names(output)
```

---

### Low-dimensional representation of subjects

`A_hat` is the joint score matrix: each row places one subject in the shared
r-dimensional space. Read it like PCA scores.

```{r plot_subject_loading, message=FALSE, warning=FALSE}
plot_subject_loading(output$A_hat, group = group_subID)
```


---

### Plot the temporal loading functions

`Zeta_m` shows when each component is active within a modality: a peak marks the
time where that component's signal is strongest.

```{r plot_time_loading, message=FALSE, warning=FALSE}
plot_time_loading(output) +
  labs(x = "Minutes from baseline", title = "Temporal loadings by modality")
```

---

### Plot the feature loadings

Feature loadings rank each modality's features by their contribution to a
component. Below: top 1% by absolute loading, negative in red, positive in blue.

```{r plot_feature_loading, message=FALSE, warning=FALSE}
plots_feat <- plot_feature_loading(output, pct = 0.05)
meta_feat <- plot_feature_loading(output, pct = 0.01) # top 1% so not as busy

# display one modality at a time:
plots_feat$cytokine
meta_feat$metabolome
plots_feat$lipid
plots_feat$protein
```

---

### Trajectories of individual features

Use the loading rankings to pick features worth inspecting directly. Here, two
top cytokines (GLP-1, Insulin) smoothed over time and split by sex.

```{r plot_feature_summary, message=FALSE, warning=FALSE}
feat_mat <- ipop[["cytokine"]][, c("GLP1", "INSULIN")]

plot_feature_summary(
  feature_mat = feat_mat,
  time_vec    = timepoints[[1]],
  group_vec   = group_sex,
  bws         = 10
) + labs(x = "Minutes from baseline")
```

---

### Subject trajectories (meta-features)

Loadings act as weights to collapse a modality into one "meta-feature"
trajectory per subject and component — handy for group comparisons.
`metafeature_aggregate` uses observed data; `_est` uses the de-noised fit.

```{r plot_metafeature, message=FALSE, warning=FALSE}
# returns a named list of ggplots, one per modality
traj_plots <- plot_metafeature(output$metafeature_aggregate, group = group_subID)

# display one modality:
traj_plots$cytokine + labs(x = "Minutes from baseline")
```

---


### Modality loading correlations

Features with similar loading profiles co-respond. Kendall correlations between
loading vectors expose these modules — we refit with more components (`r` = 10)
for a richer profile.

```{r run_multitempted2}
output2 <- multitempted_all(
  featuretables = featuretables,
  timepoints    = timepoints,
  subjectID     = subjectID,
  transforms    = "none", # data are already log10-transformed
  do_ratio      = FALSE,  # not raw counts
  r             = 10  # computing more components to compute correlations
)
```

```{r corrplot_within, message=FALSE, warning=FALSE}
# extract feature loadings per modality (transpose to r x p for cor())
cyto_loadings  <- t(as.matrix(output2$B_hat[["cytokine"]]))
metab_loadings <- t(as.matrix(output2$B_hat[["metabolome"]]))
lipid_loadings <- t(as.matrix(output2$B_hat[["lipid"]]))
prot_loadings  <- t(as.matrix(output2$B_hat[["protein"]]))

# within-modality feature correlation matrices
cyto_corr_mat  <- cor(cyto_loadings,  method = "spearman")
metab_corr_mat <- cor(metab_loadings, method = "spearman")
lipid_corr_mat <- cor(lipid_loadings, method = "spearman")
prot_corr_mat  <- cor(prot_loadings,  method = "spearman")

corrplot(cyto_corr_mat,  method = "color", tl.cex = 0.001, order = "hclust", 
         mar = c(0, 0, 2, 0), title = "Cytokine Loadings Spearman Corr Heatmap")
corrplot(metab_corr_mat, method = "color", tl.cex = 0.001, order = "hclust", 
         mar = c(0, 0, 2, 0), title = "Metabolite Loadings Spearman Corr Heatmap")
corrplot(lipid_corr_mat, method = "color", tl.cex = 0.001, order = "hclust", 
         mar = c(0, 0, 2, 0), title = "Lipid Loadings Spearman Corr Heatmap")
corrplot(prot_corr_mat,  method = "color", tl.cex = 0.001, order = "hclust", 
         mar = c(0, 0, 2, 0), title = "Protein Loadings Spearman Corr Heatmap")
```

Cross-modality correlations surface features from different modalities driven by
the same components (rectangular ordered heatmaps via `pheatmap`).

```{r corrplot_cross, message=FALSE, warning=FALSE}
cyto_v_metab <- cor(cyto_loadings, metab_loadings, method = "spearman")
cyto_v_lipid <- cor(cyto_loadings, lipid_loadings, method = "spearman")
cyto_v_prot  <- cor(cyto_loadings, prot_loadings,  method = "spearman")
metab_v_lipid <- cor(metab_loadings, lipid_loadings, method = "spearman")
metab_v_prot  <- cor(metab_loadings, prot_loadings,  method = "spearman")
lipid_v_prot  <- cor(lipid_loadings, prot_loadings,  method = "spearman")

pheatmap(cyto_v_metab,
         clustering_method = "complete",
         color = colorRampPalette(c("red", "white", "blue"))(50),
         main = "Cytokine (row) vs Metabolome (col) Loadings Corr Heatmap",
         fontsize_row = 0.001,
         fontsize_col = 0.001)
pheatmap(cyto_v_lipid,
         clustering_method = "complete",
         color = colorRampPalette(c("red", "white", "blue"))(50),
         main = "Cytokine (row) vs Lipid (col) Loadings Corr Heatmap",
         fontsize_row = 0.001,
         fontsize_col = 0.001)
pheatmap(cyto_v_prot,
         clustering_method = "complete",
         color = colorRampPalette(c("red", "white", "blue"))(50),
         main = "Cytokine (row) vs Protein (col) Loadings Corr Heatmap",
         fontsize_row = 0.001,
         fontsize_col = 0.001)
pheatmap(metab_v_lipid,
         clustering_method = "complete",
         color = colorRampPalette(c("red", "white", "blue"))(50),
         main = "Metabolome (row) vs Lipid (col) Loadings Corr Heatmap",
         fontsize_row = 0.001,
         fontsize_col = 0.001)
pheatmap(metab_v_prot,
         clustering_method = "complete",
         color = colorRampPalette(c("red", "white", "blue"))(50),
         main = "Metabolome (row) vs Protein (col) Loadings Corr Heatmap",
         fontsize_row = 0.001,
         fontsize_col = 0.001)
pheatmap(lipid_v_prot,
         clustering_method = "complete",
         color = colorRampPalette(c("red", "white", "blue"))(50),
         main = "Lipid (row) vs Protein (col) Loadings Corr Heatmap",
         fontsize_row = 0.001,
         fontsize_col = 0.001)
```

---

### Correlation network

Thresholding a correlation matrix gives a co-response network: nodes are
features, edges join strongly correlated pairs (blue = positive; red = negative).

```{r corr_network, message=FALSE, warning=FALSE}
# choose a modality correlation matrix and threshold,

choose_corr <- metab_corr_mat # change "metab" to one of "cyto", "lipid", or "prot"

threshold  <- 0.85 # change threshold if too sparse or too dense
adjacency_matrix <- ifelse(abs(choose_corr) >= threshold, choose_corr, 0)
diag(adjacency_matrix) <- 0

g <- graph_from_adjacency_matrix(
  adjacency_matrix,
  mode     = "undirected",
  weighted = TRUE,
  diag     = FALSE
)

plot(
  g,
  vertex.size  = 2,
  vertex.label = NA,
  edge.width   = abs(E(g)$weight) * 2,
  edge.color   = ifelse(E(g)$weight > 0, "steelblue", "tomato"),
  layout       = layout_with_fr(g, weights = abs(E(g)$weight))
)
```

---

## Development

*Contributors only — not needed to use the package.* Work against the source
tree with `devtools`:

```{r dev, eval=FALSE}
# from the package root
devtools::load_all()   # load the in-development source
devtools::test()       # run the test suite
devtools::document()   # regenerate man pages from roxygen comments
devtools::build_readme()  # re-render this README.md from README.Rmd
```

A Docker image (rocker/RStudio, amd64) is provided for a reproducible environment:

```{zsh, eval=FALSE}
docker build . --platform=linux/amd64 -t multi.tempted
docker run --platform=linux/amd64 \
  -v $(pwd):/home/rstudio/work -e PASSWORD=123 -p 8787:8787 multi.tempted
# open http://localhost:8787 (user: rstudio, password: 123),
# then open multi.tempted.Rproj
```
